Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-21 paper.tex
Update S3 benchmark text

One of the key benefits of Zarr is its cloud-native design, 
but it also works well on standard file systems, where 
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
benchmarks in the following sections before presenting cloud-based benchmarks. 
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
on compression performance, which uses a custom file format 
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data 
from subsets of a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so it provides 
a reasonable and easily reproducible data point 
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic 
(see the Genomics England case-study
for benchmarks on a large human dataset that includes 
many such fields). 
Note, however, that such minimal, genotype-only data 
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

@@

The discrepancy between elapsed time and total CPU time for Zarr 
benchmarks in
Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
and~\ref{fig-whole-matrix-decode} are mostly due to the overhead
of accessing a large number of small chunk files and the reduced
effectiveness of file-system read-ahead caching.
While having many independent chunks maps 
naturally to high-performance parallel access patterns
in the cloud store setting~\citep{durner2023exploiting} 
(see also the section on accessing Genomics England data stored on S3), 
such file access patterns can be problematic in the 
classical file-system setting.
The Zarr Zip store (which is simply a Zip archive of the standard directory 
hierarchy) provides a useful alternative here and can 
significantly reduce overall processing time. To illustrate this
we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
which we measure the time required to decode the entire genotype
matrix (thus serving as a lower-bound for any subsequent computations).
In the case of $10^6$ samples, the Zip file requires slightly more 
% Full data table above ^
% 58         48129895      1000000    7254858                zarr    23.752762
% 59         48129895      1000000    7254858            zarr.zip    24.240338
storage space than the directory hierarchy (23.75GiB and 24.24G
for file system and Zip file, respectively)
and while total 
CPU time is similar (193m vs 187m), the elapsed wall clock
time for the Zip file version is significantly less 
(258m vs 188m).
        % tool  user_time  sys_time    cpu_time   wall_time
% 11      zarr   11383.82    191.01  192.913833  258.598778
% 29  zarr.zip   11198.37     43.13  187.358333  188.659252

@@

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step process. We 
first converted the 106 VCF files (12.81 TiB) for chromosome 2
into the intermediate columnar format (ICF). This task was 
split into 14,605 partitions, and distributed using the Genomics England
HPC cluster. The average run-time per partition was 20.7 min.
The ICF representation used a total
of 9.94 TiB over 3,960,177 data storage files. 
We then converted the ICF to Zarr, partitioned into
5989 independent jobs, with an 18.6 min average run time.
This produced a dataset with 44 arrays, consuming a 
total of 2.54 TiB of storage over 6,312,488
chunk files. This is a roughly 5X reduction in total storage 
space over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.
We do not compare with other tools such as Genozip and 
Savvy here because  they have fundamental limitations 
(as shown in earlier simulation-based benchmarks),
and conversion of these large VCFs is a major undertaking.

@@

Finally, to explore the use of cloud computing services with VCF Zarr,
the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
Using a single-threaded process on an EC2 instance the benchmarks above were
repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
HPC system above.

This single-threaded approach is not optimal for cloud computing, as to optimise
both runtime and costs parallel processing is required. Using 64 parallel processes
on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
The network data rate was 127 MB/s, far below the capbility of the instance,
indicating that this process was compute-constrained. When calculating af-dist after
decompression the throughput was 24.6 GB/s. With this parallel approach the 
af-dist calculation above was then completed in 4.12s elapsed time.
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: $4.08 * 1.27
% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.)

@@

The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
the from the 245,388 exome-like samples from chromosome 20
using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
10,000 samples.
For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-all-of-us-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[h]
\includegraphics{figures/subset-matrix-compute-supplemental}
\caption{Compute performance on a large subset of the genotype matrix.
Total CPU time required to run the af-dist calculation for
a subset of half of the samples and 10000 variants from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
Elapsed time is also reported (dotted line). 
Genozip did not run for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute. 
\label{fig-subset-matrix-compute-supplemental}}
\end{figure}
\begin{figure}[h]

\includegraphics{figures/compression-compressor}
\caption{Effects of Blosc compression codec on compression ratio on call-level 
fields in 1000 Genomes data. 
In all cases compression level=7 was used, with a variant
chunk size of 10,000 and sample chunk size of 1,000.
Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
\label{fig-compression-compressor}}
\end{figure}

>>>>

One of the key benefits of Zarr is its cloud-native design, 
but it also works well on standard file systems, where 
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported,
and has certain advantages as discussed later).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
benchmarks in the following sections before presenting cloud-based benchmarks. 
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
on compression performance, which uses a custom file format 
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data 
from subsets of a large and highly realistic 
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture 
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset 
have been shown to closely follow 
observations~\cite{anderson2023on} and so it provides 
a reasonable and easily reproducible data point 
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic 
(see the Genomics England case-study
for benchmarks on a large human dataset that includes 
many such fields). 
Note, however, that such minimal, genotype-only data 
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

@@

The discrepancy between elapsed time and total CPU time for Zarr 
benchmarks in
Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
of accessing a large number of small chunk files and 
consequent reduced effectiveness of file-system read-ahead caching.
While having many independent chunks maps 
naturally to high-performance parallel access patterns
in the cloud store setting~\citep{durner2023exploiting} 
(see also the section on accessing Genomics England data stored on S3), 
such file access patterns can be problematic in the 
classical file-system setting.
The Zarr Zip store (which is simply a Zip archive of the standard directory 
hierarchy) provides a useful alternative here and can 
significantly reduce overall processing time. To illustrate this
we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
which we measure the time required to decode the entire genotype
matrix (thus serving as a lower-bound for any subsequent computations).
In the case of $10^6$ samples, the Zip file requires slightly more 
% Full data table above ^
% 58         48129895      1000000    7254858                zarr    23.752762
% 59         48129895      1000000    7254858            zarr.zip    24.240338
storage space than the directory hierarchy (23.75GiB and 24.24G
for file system and Zip file, respectively)
and while total 
CPU time is similar (193m vs 187m), the elapsed wall clock
time for the Zip file version is significantly less 
(258m vs 188m).
        % tool  user_time  sys_time    cpu_time   wall_time
% 11      zarr   11383.82    191.01  192.913833  258.598778
% 29  zarr.zip   11198.37     43.13  187.358333  188.659252

@@

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step process. We 
first converted the 106 VCF files (12.81 TiB) for chromosome 2
into the intermediate columnar format (ICF). This task was 
split into 14,605 partitions, and distributed using the Genomics England
HPC cluster. The average run-time per partition was 20.7 min.
The ICF representation used a total
of 9.94 TiB over 3,960,177 data storage files. 
We then converted the ICF to Zarr, partitioned into
5989 independent jobs, with an 18.6 min average run time.
This produced a dataset with 44 arrays, consuming a 
total of 2.54 TiB of storage over 6,312,488
chunk files (using a chunk size of $10^4$ variants $\times 10^3$ 
samples). This is a roughly 5X reduction in total storage 
space over the original VCF. 
The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.
We do not compare with other tools such as Genozip and 
Savvy here because  they have fundamental limitations 
(as shown in earlier simulation-based benchmarks),
and conversion of these large VCFs is a major undertaking.

@@

Finally, to explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3 
standard bucket. We repeated the benchmarks above the same subset
562,640 variants, and the same 
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed. 
Creating the filter mask
required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained from the 
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it 
is important to have many requests running concurrently to achieve 
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes 
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s 
occurs at 64 processes, with a corresponding 
network data rate of 
% FIXME MiB/s
127 MB/s, far below the capbility of the instance,
indicating that this process was compute-constrained.

When calculating af-dist after
decompression the throughput was 24.6 GB/s. With this parallel approach the 
af-dist calculation above was then completed in 4.12s elapsed time.
This level of throughput would complete af-dist for all chromosomes for all samples
% compute: $4.08 * 1.27
% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.)

@@

The compression was performed on the All of Us Researcher Workbench platform 
(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
232,809 chunk files.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[h]
\includegraphics{figures/subset-matrix-compute-supplemental}
\caption{Compute performance on a large subset of the genotype matrix.
Total CPU time required to run the af-dist calculation for
a subset of half of the samples and 10000 variants from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
Elapsed time is also reported (dotted line). 
Genozip did not run for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute. 
\label{fig-subset-matrix-compute-supplemental}}
\end{figure}

\begin{figure}[h]
\includegraphics{figures/s3-throughput}
\caption{Scalability of genotype data processing on AWS.
\label{fig-s3-throughput}}
\end{figure}

\begin{figure}[h]
\includegraphics{figures/compression-compressor}
\caption{Effects of Blosc compression codec on compression ratio on call-level 
fields in 1000 Genomes data. 
In all cases compression level=7 was used, with a variant
chunk size of 10,000 and sample chunk size of 1,000.
Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
\label{fig-compression-compressor}}
\end{figure}
